---
title: "R Linear Models"
author: "John Minter"
date: "Started: 2018-09-07, Last modified: 2018-10-11"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
library(knitr)
library(ggplot2)
options(width = 72)
opts_chunk$set(message = F, error = F, warning = F, comment = NA,
               fig.align = 'center', dpi = 100, tidy = F,
               cache.path = '.cache/', fig.path = 'figure/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

[Back to Index](../README.html)


# Fitting a linear model to DTSA-II simulation data

Load the data

```{r loadDtsaData}
load('./dat/dt2_c_on_si_7kv.rda')
print(dt2_c_on_si_7kv)
```

Set constants

```{r setConstants}
e0 <-7
```

Plot the C data


```{r plotDTSAcK}
plt_ti <- sprintf("DTSA-II simulation of C on Si at %g kV", e0)
plt <- ggplot(dt2_c_on_si_7kv) +
       geom_point(aes(x = t.C.nm, y = k.C.mu), color='blue', size=2) +
       geom_smooth(aes(x = t.C.nm, y = k.C.mu), color='red') +
       xlab(label="C thickness [nm]") +
       ylab(label="kC") +
       ggtitle(plt_ti) +
       theme(axis.text=element_text(size=12),
             axis.title=element_text(size=12),
             # center the title
             plot.title = element_text(hjust = 0.5))
print(plt)
```

Plot the Si data

```{r plotDTSAsiK}
plt_ti <- sprintf("DTSA-II simulation of C on Si at %g kV", e0)
plt <- ggplot(dt2_c_on_si_7kv) +
       geom_point(aes(x = t.C.nm, y = k.Si.mu), color='blue', size=2) +
       geom_smooth(aes(x = t.C.nm, y = k.Si.mu), color='red') +
       xlab(label="C thickness [nm]") +
       ylab(label="kSi") +
       ggtitle(plt_ti) +
       theme(axis.text=element_text(size=12),
             axis.title=element_text(size=12),
             # center the title
             plot.title = element_text(hjust = 0.5))
print(plt)
```


These plots look pretty linear. Let's fit linear models with constrained
intercepts. In the case of the kC, **we force the intercept to be zero.**
In the case of kSi,
**we set an offset of 1.0 and set the model to zero intercept**.
This means **we will add 1.0 to the negative slope** output by
the linear model. Hat Tip to
[Michael Bedward](https://stat.ethz.ch/pipermail/r-help/2010-December/262126.html)
for the correct way to do this in R.

```{r linMod}

lin.mod.k.C  <- lm(k.C.mu  ~ 0 + t.C.nm, data = dt2_c_on_si_7kv)
lin.mod.k.Si <- lm(k.Si.mu ~ 0 + t.C.nm,
                   offset=rep(1.0, length(t.C.nm)),
                   data = dt2_c_on_si_7kv)
# print(mod$coefficients)
print(lin.mod.k.C)
print(lin.mod.k.Si)
```

[Back to Index](../README.html)
