---
title: "The microscopist's use cases for ggplot2"
author: "J. R. Minter"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    csl: ../inc/american-chemical-society-with-titles-sentence-case-doi.csl
    css: ../theme/jm-blue-vignette.css
    toc: true
    number_sections: yes
bibliography: ../inc/tips.bib
---

[Back to Index](../README.html)

I typically have spurts where I use  R and ggplot2 [@Wickham2009a] to 
analyze data and find myself saying, "I've done this before, but where
is the code?" These exemplars should make it easy to get back up to speed
quickly. As Karl Broman noted: 

> Your closest collaborator is you six months ago and you don't respond to email!

So this is for "future me" and anybody else who needs it. See also the
**ggplotExamples** project in your private jrm-solutions repo.


First, let's load the packages we need.

```{r}
library(ggplot2)
library(latex2exp)
```


# Plot/fit data from processing EDS spectra

In this case the spectra were simulated using DTSA-II [@Ritchie2011b].

## Step 1: Load the data

```{r}
fi <-'./dat/csv/C-ctd-Si-5-kV-10000-Traj.csv'
df <- read.csv(fi, header=TRUE, as.is=TRUE)
print(head(df))
```

## Compute the C/Si peak ratio and uncertainty

```{r}
df$c.to.si.mu <- df$C.Int.mu / df$Si.Int.mu
df$c.to.si.unc <- sqrt((df$C.Int.unc/df$C.Int.mu)**2 + 
                       (df$Si.Int.unc/df$Si.Int.mu)**2)
```

## Make our plot

```{r}
ctosiInt <- ggplot(df, aes(x=t.C.nm, y=c.to.si.mu)) + 
    geom_errorbar(aes(ymin=c.to.si.mu - 1.96*c.to.si.unc,
                      ymax=c.to.si.mu + 1.96*c.to.si.unc), width=.1) +
    stat_smooth(method = "loess") +
    annotate("text", label = 'loess smooth',
             x = 70, y = 1.1,
             size = 5, colour = "blue") +
    geom_point(color='black', size=2) +
    xlab("C thickness [nm]") +
    ylab("C-K/Si-K X-ray intensity ratio") +
    ggtitle("Monte Carlo Model of C-coatings on Si at 5 kV") +
    theme(axis.text=element_text(size=12),
          axis.title=element_text(size=14),
          plot.title = element_text(hjust = 0.5)) # center the title

print(ctosiInt)
```

## Step 2: Compute LOESS interpolated predicted values

```{r}
t.C.nm <- seq(from=0.5, to=100, by=0.5)
df2 <- data.frame(t.C.nm=t.C.nm, c.to.si.mu=t.C.nm)
df2$c.to.si.mu = predict(loess(c.to.si.mu~t.C.nm, data=df2))
```

## Step 3: Write some output

```{r}

fi <- './dat/pdf/c-ctd-si-series-plt-5kV.pdf'
ggsave(ctosiInt, file=fi, width=9.0, height=6.0, units="in", dpi=300)

fi <- './dat/png/c-ctd-si-series-plt-5kV.png'

# we want the png to be close to 1024x768...
ggsave(ctosiInt, file=fi, width=9.0, height=6.0, units="in", dpi=113.7778)
fi <- './dat/csv/c-ctd-si-loess-pred-5kV.csv'
write.csv(df2, file=fi, row.names=FALSE)

```

# Plotting PhiRhoZ curves from DTSA-II Monte Carlo simulations

DTSA-II computes the PhiRhoZ curves proposed by [@Packwood1981a].

## Step 1: Load the data.

Use the helper function from package **rEDS** to load the data and explore
the columns. This needs version **0.0.1.1** or later. Note: the **PhiRhoZ.csv**
file outputted by DTSA-II is really **tab-delimited** and has two header
lines (the headers we want and a second units line.) The **rEDS** package
contains a wrapper function (**prepDataframeDtsaPhiRhoZ**) that cleans
the data up and generates a dataframe from which we can choose the 
features that we really want to plot.

```{r, doRead,message=FALSE}
library(rEDS)

fi <- './dat/csv/PhiRhoZ.csv'
df <- prepDataframeDtsaPhiRhoZ(fi)
print(head(df))    
```


## Step 2: Make a dataframe with the transitions we care about.

While we are at it, let's change the dataframe headers to make
it easier to make the plots... We are plotting the generated
and emitted X-rays. Let's make simple mnemonics... 

```{r makeVecs}
                     
z <- df[,1]
okgen <- df[,8]
okemi  <- df[,9]
sikgen <- df[,12]
sikemi  <- df[,13]

df2 <- data.frame(Z.um=z,
                 O.Gen=okgen, O.Emi=okemi,
                 Si.Gen=sikgen, Si.Emi=sikemi)
print(head(df2))

```

Isn't that more readable! 

In this case we also want to remove the first line because it is the thin
C coating that creates a discontinuity. If we remove this first line,
the smooth works better...

```{r removeStep, message=FALSE}
df3 <- df2[-1, ]

print(head(df3))
```

## Step 3: Plot the Si data.

Note the use of the Unicode in the **xlab** string to create the mu...


```{r plotSi, message=FALSE}
library(ggplot2)
plt <- ggplot(df3, aes(x = Z.um)) +
              geom_point(aes(y = Si.Gen), colour="blue") +
              geom_smooth(aes(x= Z.um, y=Si.Gen),
                          span = 0.18, color = 'blue') +
              annotate("text", label = 'Generated Si K-L3',
                       x = .8, y = 2.25,
                       size = 5, colour = "blue") +
              geom_point(aes(y = Si.Emi), colour="red") +
              geom_smooth(aes(x= Z.um, y=Si.Emi),
                          span = 0.1, color='red') +
              annotate("text", label = 'Emitted',
                       x = .3, y = .75,
                       size = 5, colour = "red") +
              ylab(label="Si Intensity") +
              # note the Unicode string for mu
              xlab("depth [\U00B5m]") +
              ggtitle("DTSA-II Monte Carlo PhiRhoZ Model 10 nm C on SiO2 at 10 kV") +
              theme(axis.text=element_text(size=12),
                    axis.title=element_text(size=14),
                    plot.title = element_text(hjust = 0.5)) # center the title

print(plt)

```

## Step 4: Plot the O data.


```{r plotO, message=FALSE}
library(ggplot2)

df3 <- df2[-1, ]

plt <- ggplot(df3, aes(x = Z.um)) +
              geom_point(aes(y = O.Gen), colour="blue") +
              geom_smooth(aes(x= Z.um, y=O.Gen),
                          span = 0.18, color = 'blue') +
              annotate("text", label = 'Generated O K-L3',
                       x = .8, y = 2.25,
                       size = 5, colour = "blue") +
              geom_point(aes(y = O.Emi), colour="red") +
              geom_smooth(aes(x= Z.um, y=O.Emi),
                          span = 0.1, color='red') +
              annotate("text", label = 'Emitted',
                       x = .3, y = .75,
                       size = 5, colour = "red") +
              ylab(label="O Intensity") + 
              # note the Unicode string for mu
              xlab("depth [\U00B5m]") +
              ggtitle("DTSA-II Monte Carlo PhiRhoZ Model 10 nm C on SiO2 at 10 kV") +
              theme(axis.text=element_text(size=12),
                    axis.title=element_text(size=14),
                    plot.title = element_text(hjust = 0.5)) # center the title

print(plt)

```



# Analyze particle size and shape data

## Step 1: Load the data

Load the data from a **.csv** file and decide if we want to save plots.

```{r, results='asis'}
grains <- read.csv("./dat/csv/grains.csv", header = TRUE, as.is=TRUE)
print(names(grains))
bSave <- FALSE
```

## Step 2: Making a Histogram With ggplot2

You have two options to make a histogram with ``ggplot2`` package. You can either use the ``qplot()`` function, which looks very much like the ``hist()`` function:

```{r}
pHistoAgxDiam <- ggplot(grains, aes(ecd.nm)) +
  geom_histogram(binwidth = 5) +
  # scale_x_continuous(breaks = seq(from = 7, to = 20, by = 1)) +
  # scale_y_discrete(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16)) +
  ggtitle("Grain Equivalent Circular Diameter") +
  theme(plot.title = element_text(lineheight=2, size=12)) +
  # labs(x=TeX("diameter ($\\mu$m)"), y="Count")
  labs(x="diameter (nm)", y="Count") 

if (bSave == TRUE) {
  ggsave("./dat/png/grain-ecd-histo.png", plot=pHistoAgxDiam,
         width=6, height=4, units="in", dpi=150)
}
print(pHistoAgxDiam)
```

## Step 3: Make a histogram for aspect ratio.


```{r}
print(summary(grains$a.r))

pHistoAgxAr <- ggplot(grains, aes(a.r)) +
  geom_histogram(bins=50) +  # binwidth = 5) +
  # scale_x_continuous(breaks = seq(from = 7, to = 20, by = 1)) +
  # scale_y_discrete(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16)) +
  ggtitle("Grain Aspect Ratio") +
  theme(plot.title = element_text(lineheight=2, size=12)) +
  # labs(x=TeX("diameter ($\\mu$m)"), y="Count")
  labs(x="diameter (nm)", y="Count") 

if (bSave == TRUE) {
  ggsave("./dat/png/grain-ecd-histo.png", plot=pHistoAgxAr,
         width=6, height=4, units="in", dpi=150)
}
print(pHistoAgxAr)
```

## Step 4: Make some classes and plot.

```{r}
equiAxial <- grains$ecd.nm[grains$a.r < 1.05]
elongated <- grains$ecd.nm[grains$a.r >= 1.05]
df1 <- data.frame(ecd=equiAxial,shape='equiaxial')
df2 <- data.frame(ecd=elongated,shape='elonganted')
shapes <- rbind(df1,df2)
rm(df1, df2)

pBoxPlotEcdAr <- ggplot(shapes, aes(x=shape, y=ecd)) +
  geom_boxplot(aes(fill = shape)) +
  # scale_y_continuous(breaks = seq(from = 10, to = 20, by = 1)) +
  # ggtitle(TeX("Beads $\\geq 10\\mu$m\n from Class 837 Support")) +
  # theme(plot.title = element_text(lineheight=2, size=12)) +
  labs("equivalent circular diameter (nm)") 

if (bSave == TRUE) {
   ggsave("./dat/png/grain-ecd-shape-boxplot.png",
          plot=pBoxPlotEcdAr, width=6, height=4,
          units="in", dpi=150)
}
print(pBoxPlotEcdAr)
```

## Step 5: Make a scatter plot.


```{r}
pScatterAgXArEcd <- ggplot(grains, aes(x=a.r, y=ecd.nm)) +
  geom_point(colour="blue") + 
  # scale_x_continuous(breaks = seq(from = 7, to = 20, by = 1)) +
  # scale_y_discrete(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16)) +
  # ggtitle("Grain Aspect Ratio") +
  # theme(plot.title = element_text(lineheight=2, size=12)) +
  # labs(x=TeX("diameter ($\\mu$m)"), y="Count")
  labs(x="aspect ratio", y="ecd [nm]") 
if (bSave == TRUE) {
  ggsave("./dat/png/grain-ecd-ar.png", plot=pScatterAgXArEcd,
         width=6, height=4, units="in", dpi=150)
}
print(pScatterAgXArEcd)
```

```{r}
pScatterAgXCircEcd <- ggplot(grains, aes(x=circ, y=ecd.nm)) +
  geom_point(colour="blue") + 
  # scale_x_continuous(breaks = seq(from = 7, to = 20, by = 1)) +
  # scale_y_discrete(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16)) +
  # ggtitle("Grain Aspect Ratio") +
  # theme(plot.title = element_text(lineheight=2, size=12)) +
  # labs(x=TeX("diameter ($\\mu$m)"), y="Count")
  labs(x="circularity", y="ecd [nm]") 
if (bSave == TRUE) {
  ggsave("./dat/png/grain-ecd-circ.png", plot=pScatterAgXCircEcd,
         width=6, height=4, units="in", dpi=150)
}
print(pScatterAgXCircEcd)
```

## Step 6: Plot the diameter distribution of our dominant morphology.

```{r}
dominant <- grains[grains$a.r <= 1.2, ]
dominant <- dominant[dominant$circ >= 0.8, ]

ecd.mu <- mean(dominant$ecd.nm)
ecd.sd <- sd(dominant$ecd.nm)


pDomGrainDiam <- ggplot(dominant, aes(ecd.nm)) +
  geom_histogram(binwidth = 5) +
  # scale_x_continuous(breaks = seq(from = 7, to = 20, by = 1)) +
  # scale_y_discrete(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16)) +
  ggtitle("Dominant Population Equivalent Circular Diameter") +
  theme(plot.title = element_text(lineheight=2, size=12)) +
  # labs(x=TeX("diameter ($\\mu$m)"), y="Count")
  labs(x="diameter (nm)", y="Count") 

if (bSave == TRUE) {
 ggsave("./dat/png/dom-grain-ecd-histo.png", plot=pDomGrainDiam,
        width=6, height=4, units="in", dpi=150)
}
print(pDomGrainDiam)
```

## Step 7: Make a plot of computed Gaussian data.

```{r}

ecd <- rnorm(100000, mean=ecd.mu, sd=ecd.sd)
df <- data.frame(ecd=ecd)
theoHistPlot <- ggplot(df, aes(ecd)) +
  geom_histogram(binwidth = .1) +
  labs(x="diameter (nm)", y="Count") 

print(theoHistPlot)
```

[Back to Index](../README.html)

## References