---
title: "Excel Tips"
author: "J. R. Minter"
date: "Started: 2016-09-16, Last modified: 2017-10-28"
output:
  knitr:::html_vignette:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
---
[Back to Index](../README.html)

```{r, echo=FALSE}
library(png)
library(grid)
```

# Tips from [Joel Spolsky](https://www.youtube.com/watch?v=0nbkaYsR94c)

Joel noted:

> The way you are using Excel causes errors, creates incomprehensible
spaghetti spreadsheets, and makes me want to stab out my own eyes.

## A1 vs R1C1

> You have Excel set up in Baby Mode (A1), with ABC's across the top like some
kind of children's educational song, and 123 down the left. You can also set
it up in Big Boy Mode (R1C1), with numbers across the top and down the left.
If you learn how to operate in R1C1 mode, it's easier to understand relative
vs. absolute references.

```{r fig.width=5.5, fig.height=4.485, echo=FALSE,fig.align='center', fig.cap="Set R1C1 mode for better relative references."}
img <- readPNG("png/set-r1c1-mode.png")
grid.raster(img)
```

In his example he made a formula that used the `&` for concatenation of strings,
including values from two cells and a fixed string. Note how the use of relative
references in **RC** notation makes the locations easy to understand.


```{r fig.width=5.5, fig.height=1.557, echo=FALSE,fig.align='center', fig.cap="Set R1C1 mode for better relative reference with concatenation operator."}
img <- readPNG("png/concatenation-rel-refs.png")
grid.raster(img)
```

Joel makes frequent use of **Keyboard Shortcuts** -- such as **Ctl-C** for
**copy**. He notes that most of the time when cleaning data scraped from
the web (or other spreadsheets), you do **not** want to use **paste**, but
instead **past values** because Excel gets the references wrong...

Joel notes:

> Every cell has a **formula**, a **value**, and **formatting**. It shows you
the value. When you are cutting and pasting data from external sources, you
generally just want the values.

## Absolute vs. Relative reference

### A1 mode

```
$ means absolute

A1   - both the column and the row are relative
$A1  - the column is absolute
A$1  - the row is absolute
$A$1 - both the column and the row are absolute
```

### R1C1 mode

```
[] means relative

R1C1   - absolute position
RC[-1] - same row, one column to the left
R[3]C2 - absolute column, three rows down
RC     - this cell
```

### Quick toggles

1. There is a quick toggle from absolute and relative references. On Windows it
is **F4**. On Mac OSX it is **Cmd+T**.
2. **Cntrl+`** (backtick) - a toggle between displaying Formulas and Values

## Data Entry shortcuts

### Getting the right column width

1. Double click between column headers to adjust to fit widest column
2. Select multiple columns first to do this to all columns
3. To make many columns the same width, select multiple columns first and
adjust width

### Ctrl-Enter

1. **Enter** - enters your changes in active cell
2. **Ctrl+Enter** - enters your changes in the entire selection

### Fill Down/Fill Right

1. **Ctrl+D** - fill down. Can also "pull down" the formula from the cell above
2. **Ctrl+R** - fill right.

### The handy fill-handle

That little fill handle does a million things:

- **Fill down** by dragging it
- **Ride The Range** by double clicking it (riding the range is filling down until the cell to the left is empty)
- **Give it a pattern** and it will continue it (Mon/Tue, Jan/Feb, 1/2/3, etc)
- **Drag backwards** to erase
- **Right-Drag** for linear and growth trends

### Copy | Paste values

- Cells have formulas and values.

- If you don't need the formula any more, Copy | Paste Values.

- This is also a good way to get rid of formatting. Remember: Formatting is for the weak.

Other useful things are hiding behind the **Paste Special** menu. The most popular are add/subtract/multiply/divide, transpose, and skip blanks.


```{r fig.width=4.5, fig.height=3.7, echo=FALSE,fig.align='center', fig.cap="The **Paste Special** menu."}
img <- readPNG("png/paste-special.png")
grid.raster(img)
```

### Paint Formats

Learn how that dumb paintbrush works. You create a cell formatted the way you like, highlight it, click the paintbrush tool, and drag over your cells. This is especially helpful for tables...

### Cntrl+A (Cmd+A MacOSX)

Stop trying to select while scrolling. Cmd+A (Ctrl+A on Windows) selects the
current region which means the largest island surrounded by white space.

### Text values

If your numbers are really codes (like ZIP codes) start them with apostrophe (') to prevent leading zeros from disappearing. Or use the TEXT number format.

## Clean up your formats

Seriously? I never want to see absolute or relative references. References are confusing, apt to screw up when you move things (or insert (or delete)), and the source of all bugs.

**Tables should be surrounded by whitespace**

**Always use named ranges**

See also the **Create Names** feature to do this easily.

In Excel2016 see the **Name Manager** under the **Formulas** menu.
One may use this to change the ranges.

Intersect with a named range

If Foo is a name that refers to a whole column, the formula =Foo*2 gets a single value out of Foo that is in the same row as the formula and multiplies that by 2. This is important because you can always name your columns

1. Name your columns
2. Use formulas that intersect columms
3. Just use Tables
4. Create intermediate coumns and hide them
5. Use ROUND() for values with cents
6. Use **INDEX** and **MATCH**,  *never* **VLOOKUP**
7. Why do you have **numbers** in your formulas? Create parameter entry cells and use them in formulas.

[Back to Index](../README.html)
