---
title: "Make gg_panel_plot"
author: "J. R. Minter"
date: "Started: 2019-05-25, Last modified: 2019-05-25"
output:
  html_document:
    keep_md: yes
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

[Back to ImageJ](ImageJ.html)


```{r setup, cache = F, echo = T, message = F, warning = F, tidy = F}
options(width = 72)
bSave = FALSE # don't save plots until I', satisfied...
```

I'll need
[this](https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html)
later...

```{r load_packages, message=FALSE, comment=NA}
library(knitr)
library(dplyr)
library(ggplot2)
```

# Read in the data with R

```{r readRawData, message=FALSE, comment=NA}
fi <-'./csv/pol4455-particle_sizer.csv'
df <- read.csv(fi, header=TRUE, as.is=TRUE)
df <- as_tibble(df)
df
```

Make a tibble that includes the traditional definition of circularity.

```{r makeTibble, message=FALSE, comment=NA}
area = df$Area
peri = df$Peri.
trad_circ = 4.0*pi*area/peri^2
# summary(trad_circ)

particles <- as_tibble(data.frame(lab=df$Label,
                                  ecd=df$Area.equivalent.circle.diameter,
                                  ar=df$Aspect.Ratio,
                                  circ=trad_circ,
                                  elong=df$Elong.,
                                  conv=df$Convexity))
summary(particles)
```
# The measure of circularity

I am **really** confused by the `circularity` value reported by ParticleSizer.
I noticed that all the values are close to 12. I am used to the definition of
circularity as defined by [Wikipedia](https://en.wikipedia.org/wiki/Shape_factor_(image_analysis_and_microscopy)#Circularity)
as

$\frac{4  \pi A}{P^2}$

that ranges from 0 to 1.0. So that is what we will use.

# Plot the histogram of ECD

```{r plotECD, message=FALSE, comment=NA}

#' Plot the ECD
#'
#' Make a ggplot2 plot of the ECD
#'
#' @param tib tibble - Containing the ecd
#' @param bin_width number - The width of the bin for the diameter histogram
#' @param title String - A title for the plot
#' @param xlab String - the label for the x axis, e.g. "diameter [nm]"
#' @param ylab String - the y label. Default = "Counts"
#' @param base_txt_pts String - the size in points. Default 12.
#'
#' @return plt
#'
#' @examples
#' library(rAnaLab)
#' x <- sprintf("H4100 800X: X= %.3f, %.3f",  calc.adda.mm.per.px(150.92),
#'      calc.adda.mm.per.px(148.77))
#' print(x)
#'
#'
#'
#' @export
#'
#'

plot_ecd <- function(tib, bin_width, title, xlab, ylab = "Counts",
                     base_txt_pts=12){
  plt <- ggplot(tib, aes(ecd)) +
         geom_histogram(binwidth=bin_width) +
         ggtitle(title) +
         theme(plot.title = element_text(lineheight=2, size=base_txt_pts+2)) +
         labs(x=xlab, y=ylab) +
         ggtitle(title) +
         theme(axis.text=element_text(size=base_txt_pts),
               axis.title=element_text(size=base_txt_pts+2),
              plot.title = element_text(hjust = 0.5)) +
         NULL
        return(plt)
}


plt <- plot_ecd(particles, 1.0, "Diameter distribution", "diameter [nm]", "Counts")

print(plt)
```

```{r plot_trad_circ, message=FALSE, comment=NA}

#' Plot the traditional circularity distribution
#'
#' Make a ggplot2 plot of the traditional circularity distribution
#'
#' @param tib tibble - Containing the circularity
#' @param bin_width number - The width of the bin for the circularity histogram
#' @param title String - A title for the plot
#' @param xlab String - the label for the x axis, e.g. "diameter [nm]"
#' @param ylab String - the y label. Default = "Counts"
#' @param base_txt_pts String - the size in points. Default 12.
#'
#' @return plt
#'
#' @examples
#' library(rAnaLab)
#' x <- sprintf("H4100 800X: X= %.3f, %.3f",  calc.adda.mm.per.px(150.92),
#'      calc.adda.mm.per.px(148.77))
#' print(x)
#'
#'
#'
#' @export
#'
#'

plot_circ <- function(tib, bin_width, title, xlab, ylab = "Counts",
                     base_txt_pts=12){
  plt <- ggplot(tib, aes(circ)) +
         geom_histogram(binwidth=bin_width) +
         ggtitle(title) +
         theme(plot.title = element_text(lineheight=2, size=base_txt_pts+2)) +
         labs(x=xlab, y=ylab) +
         ggtitle(title) +
         theme(axis.text=element_text(size=base_txt_pts),
               axis.title=element_text(size=base_txt_pts+2),
              plot.title = element_text(hjust = 0.5)) +
         NULL
        return(plt)
}


pltc <- plot_circ(particles, 0.01, "Circularity distribution", "Circularity", "Counts")

print(pltc)

```


```{r plot_diam_boxplot, message=FALSE, comment=NA}

the_plt <- ggplot(particles, aes(x="", y=ecd)) + 
           geom_boxplot() +
           coord_fixed(.1) +
           labs(y="ecd nm") +
           theme(axis.title.x=element_blank(),
                 axis.text.x=element_blank(),
                 axis.ticks.x=element_blank())

print(the_plt)
```


```{r ecd_qq_plot}
ecd_qq_plt <- ggplot(particles, aes(sample = ecd)) +
              stat_qq() +
              stat_qq_line() +
              labs(y="ECD [nm]") +
              coord_fixed(.5) +
              NULL
         
print(ecd_qq_plt)
```





````


pltc <- ggplot(particles, aes(circ)) +
        geom_histogram(binwidth = 0.1) +
        theme(plot.title = element_text(lineheight=2, size=12)) +
        labs(x="'ParticleSizer' Circularity", y="Count") +
        ggtitle("Circularity distribution of soft latex particles in vitreous ice") +
        theme(axis.text=element_text(size=12), axis.title=element_text(size=14),
              plot.title = element_text(hjust = 0.5)) +
        NULL

if (bSave == TRUE) {
  ggsave("png/pol4455-part-sizer-circ-histo.png", plot=pltc,
         width=6, height=4, units="in", dpi=150)
}
print(pltc)

```

Try a linear distribution panel plot that uses base graphics. This assunmes
a linear (not lognormal) particle size distribution from my old `rAnaLab`
package.

Sometime I need to migrate this function to ggplot2...


```{r linearDistPanelPlot, message=FALSE}
library(rAnaLab)

linear.distn.panel.plot(particles$ecd, n.brks = 25,
                        distn.lab = "ecd [nm]",
                        hist.legend = TRUE,
                        legend.loc = "topright",
                        kern.bw = "nrd0",
                        plt.median = TRUE,
                        scale.mult = 1.2)

```
```{r save_panel_plt, message=FALSE}
if (bSave == TRUE) {
  png("png/linearDistPanelPlot.png", width=1024, height=768, pointsize=24)
  linear.distn.panel.plot(particles$ecd, n.brks = 25,
                         distn.lab = "ecd [nm]",
                         hist.legend = TRUE,
                         legend.loc = "topright",
                         kern.bw = "nrd0",
                         plt.median = TRUE,
                         scale.mult = 1.2)
  dev.off()
}

```

[Back to ImageJ](ImageJ.html)
