---
title: "Make gg_panel_plot"
author: "J. R. Minter"
date: "Started: 2019-05-25, Last modified: 2019-05-27"
output:
  html_document:
    keep_md: yes
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

[Back to ImageJ](ImageJ.html)


```{r setup, cache = F, echo = T, message = F, warning = F, tidy = F}
options(width = 72)
bSave = FALSE # don't save plots until I', satisfied...
```

I'll need
[this](https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html)
later...

```{r load_packages, message=FALSE, comment=NA}
library(knitr)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

# Read in the data with R

```{r readRawData, message=FALSE, comment=NA}
fi <-'./csv/pol4455-particle_sizer.csv'

df <- read.csv(fi, header=TRUE, as.is=TRUE)
df <- as_tibble(df)
df
```

Make a tibble that includes the traditional definition of circularity. We
discovered that the `Circ.` column need to be divided by $4 \pi$.

```{r makeTibble, message=FALSE, comment=NA}
area = df$Area
peri = df$Peri.
trad_circ = 4.0*pi*area/peri^2
# summary(trad_circ)

particles <- as_tibble(data.frame(lab=df$Label,
                                  ecd=df$Area.equivalent.circle.diameter,
                                  ar=df$Aspect.Ratio,
                                  circ=df$Circ./(4.0*pi),
                                  elong=df$Elong.,
                                  conv=df$Convexity))
summary(particles)
```
# The measure of circularity

I was **really** confused by the `Circ.` value reported by ParticleSizer.
I noticed that all the values are close to 12. I am used to the definition of
circularity as defined by [Wikipedia](https://en.wikipedia.org/wiki/Shape_factor_(image_analysis_and_microscopy)#Circularity)
as

$\frac{4  \pi A}{P^2}$. It turns out that ParticleSizer did **not** divide by
$4 \pi$ so we will correct the result during data processing.

that ranges from 0 to 1.0. So that is what we will use.

# Plot the histogram of ECD

```{r plotECD, message=FALSE, comment=NA}

#' Plot the ECD
#'
#' Make a ggplot2 plot of the ECD
#'
#' @param tib tibble - Containing the ecd
#' @param bin_width number - The width of the bin for the diameter histogram
#' @param title String - A title for the plot
#' @param xlab String - the label for the x axis, e.g. "diameter [nm]"
#' @param ylab String - the y label. Default = "Counts"
#' @param base_txt_pts String - the size in points. Default 12.
#'
#' @return plt
#'
#' @examples
#' library(rAnaLab)
#' x <- sprintf("H4100 800X: X= %.3f, %.3f",  calc.adda.mm.per.px(150.92),
#'      calc.adda.mm.per.px(148.77))
#' print(x)
#'
#'
#'
#' @export
#'
#'

plot_ecd <- function(tib, bin_width, title, xlab, ylab = "Counts",
                     base_txt_pts=12){
         med <- median(tib$ecd)
  plt <- ggplot(tib, aes(ecd)) +
         geom_histogram(binwidth=bin_width) +
         ggtitle(title) +
         theme(plot.title = element_text(lineheight=2, size=base_txt_pts+2)) +
         labs(x=xlab, y=ylab) +
         ggtitle(title) +
         theme(axis.text=element_text(size=base_txt_pts),
               axis.title=element_text(size=base_txt_pts+2),
              plot.title = element_text(hjust = 0.5)) +
         geom_vline(xintercept=med,linetype=1, size=1.5, colour="blue") +
         NULL
        return(plt)
}

anot <- sprintf("Median is \n %.1f nm", median(particles$ecd)) 

ecd_plt <- plot_ecd(particles, 1.0, "Diameter distribution",
                "diameter [nm]", "Counts") +
       annotate("text", label = anot, x = 30, y = 150, size = 4,
                colour = "blue")

print(ecd_plt)
```

```{r plot_circ, message=FALSE, comment=NA}

#' Plot the corrected circularity distribution
#'
#' Make a ggplot2 plot of the traditional circularity distribution
#'
#' @param tib tibble - Containing the circularity
#' @param bin_width number - The width of the bin for the circularity histogram
#' @param title String - A title for the plot
#' @param xlab String - the label for the x axis, e.g. "diameter [nm]"
#' @param ylab String - the y label. Default = "Counts"
#' @param base_txt_pts String - the size in points. Default 12.
#'
#' @return plt
#'
#' @examples
#' library(rAnaLab)
#' x <- sprintf("H4100 800X: X= %.3f, %.3f",  calc.adda.mm.per.px(150.92),
#'      calc.adda.mm.per.px(148.77))
#' print(x)
#'
#'
#'
#' @export
#'
#'

plot_circ <- function(tib, bin_width, title, xlab, ylab = "Counts",
                     base_txt_pts=12){
  plt <- ggplot(tib, aes(circ)) +
         geom_histogram(binwidth=bin_width) +
         ggtitle(title) +
         theme(plot.title = element_text(lineheight=2, size=base_txt_pts+2)) +
         labs(x=xlab, y=ylab) +
         ggtitle(title) +
         theme(axis.text=element_text(size=base_txt_pts),
               axis.title=element_text(size=base_txt_pts+2),
               plot.title = element_text(hjust = 0.5)) +
         NULL
        return(plt)
}


pltc <- plot_circ(particles, 0.01, "Circularity distribution", "Circularity", "Counts")

print(pltc)

```


```{r plot_diam_boxplot, message=FALSE, comment=NA}

plot_diam_boxplot <- function(tib, c_fixed=0.1, title="Box Plot", ylab = "ECD [nm]",
                     base_txt_pts=12){

the_plt <- ggplot(particles, aes(x="", y=ecd)) + 
           geom_boxplot() +
           coord_fixed(c_fixed) +
           labs(y=ylab, x="") +
           ggtitle(title) +
           theme(axis.text=element_text(size=base_txt_pts),
                 axis.title=element_text(size=base_txt_pts+2),
                 plot.title = element_text(hjust = 0.5)) +
           NULL

return(the_plt)
}

box_plt <- plot_diam_boxplot(particles)
print(box_plt)

```


```{r ecd_qq_plot}
plot_ecd_qq_plot <- function(tib, c_fixed=0.5, title="QQ Plot", ylab = "ECD [nm]",
                     base_txt_pts=12){
base_txt_pts=12
plt <- ggplot(tib, aes(sample = ecd)) +
              stat_qq() +
              stat_qq_line() +
              labs(y=ylab, x="") +
              ggtitle(title) +
              coord_fixed(c_fixed) +
              theme(axis.text=element_text(size=base_txt_pts),
                 axis.title=element_text(size=base_txt_pts+2),
                 plot.title = element_text(hjust = 0.5)) +
              NULL
         
return(plt)
}

ecd_qq_plot <- plot_ecd_qq_plot(particles, c_fixed=0.5, title="QQ Plot",
                                ylab = "ECD [nm]", base_txt_pts=12)
print(ecd_qq_plot)
```





Try a linear distribution panel plot that uses base graphics. This assunmes
a linear (not lognormal) particle size distribution from my old `rAnaLab`
package.

Sometime I need to migrate this function to ggplot2...


```{r linearDistPanelPlot, message=FALSE}
library(rAnaLab)

linear.distn.panel.plot(particles$ecd, n.brks = 25,
                        distn.lab = "ecd [nm]",
                        hist.legend = TRUE,
                        legend.loc = "topright",
                        kern.bw = "nrd0",
                        plt.median = TRUE,
                        scale.mult = 1.2)

```
```{r save_panel_plt, message=FALSE}
if (bSave == TRUE) {
  png("png/linearDistPanelPlot.png", width=1024, height=768, pointsize=24)
  linear.distn.panel.plot(particles$ecd, n.brks = 25,
                         distn.lab = "ecd [nm]",
                         hist.legend = TRUE,
                         legend.loc = "topright",
                         kern.bw = "nrd0",
                         plt.median = TRUE,
                         scale.mult = 1.2)
  dev.off()
}

```

Make the panel plot

```{r panel_plot}

out <- grid.arrange(ecd_plt, box_plt, ecd_qq_plot, nrow=1 )
ggsave(file="png/pol4455_gg_panel_plot.png", out) 
```

[Back to ImageJ](ImageJ.html)
