---
title: "R data pipeline"
author: "J. R. Minter"
date: "Started: 2013-04-25, Last modified: 2019-04-25"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, comment=NA}
library(knitr)
options(width = 72)

suppressPackageStartupMessages(library("tidyverse"))
```


[Back to Index](../README.html)


# How to understand the pipe (`%>%`) operator

The `tidyverse` pulls the pipe (`%>%`) operator from the `magritter`
package. Understanding it is often difficult at first. The best way
to read a sequence is to read the operator as **and then**. Note that
RStudio defines a keyboard shortcut `Ctrl+Shift+M` (Win) and  `CMD+Shift+M`
(Mac).

The examples came from a tutorial given by Simon Jackson at the
useR2018 conference. See the
[youtube video)](https://www.youtube.com/watch?v=E-Vvg8uzcVM)
and the [github repository](https://github.com/drsimonj/tidyverse_tutorial-useR2018).


```{r readPriceData, comment=NA}
library(tidyverse)
bookings <- read_csv("dat/bookings.csv")
# properties <- read_csv("dat/properties.csv")
bookings %>%
  filter(price_per_night < 80) %>% 
  select(room_nights, review_score) %>% 
  head(2)
```

Pipes are not new. Note that the Linux/Unix shell has been piping
commands for decades...

We don't need to create multiple objects and we can read the sequence
easily.

Another query from the `bookings` data.

> Write a data pipeline to get a tibble of the `property_id` and 
> `status` of all bookings made on a Wednesday.

```{r checkinWedData, comment=NA}
bookings %>%
  filter(checkin_day=='wed') %>% 
  select(property_id, status) %>% 
  head(5)
```



[Back to Index](../README.html)
