---
title: "Monte Carlo Techniques in R"
date: "Started: 2017-04-27, Last modified: 2018-10-11"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---
[Back to Index](../README.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

From [here](https://blog.alookanalytics.com/2017/04/26/monte-carlo-method-in-r/)

# Analyze Stocks

Worst-case scenario for future stock price

Let’s see on a simple example how easy is to perform Monte Carlo method in R. Assume we want to calculate the worst-case scenario of a future stock price. This problem called [value at risk](https://en.wikipedia.org/wiki/Value_at_risk) is heavily used in risk management. By “worst-case scenario” we mean the value that the stock price will exceed with 99% probability (i.e., there is only 1% probability that the stock price will be below). The current price of our stock is 100 $. We want to see the possible future prices after 20 trading days.

As stated above, we need to make some assumptions about the future stock price. So let’s say that the drift over the 20 trading days is 10% (i.e. in average the price will go up to 110) and the volatility is 20%.

Second step is to generate some random inputs, which in our case means to model the future price given the current price, drift and the volatility. For our purpose we do not need anything fancy so we will use the standard stock price model called [Geometric Brownian Motion](https://en.wikipedia.org/wiki/Geometric_Brownian_motion):

S~t+1~ = S~t~ * (1 + μΔt + σε√Δt)

Where S~t~ is price in time t, μ is our drift, Δt means one period (in our case 1 trading day, i.e. one twentieth of the period), σ is our volatility and ε is a random number between 0 and 1. The core idea of Monte Carlo method is to generate the future price (which is random) high number of times to simulate what are all the situations that can occur.

In the third step we calculate the stock price at the end of 20th day given the model for all the randomly generated daily returns. This step (and also the previous ones) is demonstrated by the following code.

## define a stock price function

```{r, defineStockFunction}

#' Stock price calculation
#' 
#' Calculates stock price after n periods using standard stock price model
#' @param stock_price original stock price
#' @param n number of periods
#' @param stock_mu expected percentual stock drift over n periods
#' @param stock_sigma expecter percentual stock volatility
#' @return stock price after n periods
f_stock_return <- function(stock_price, n, stock_mu, stock_sigma){
  delta_t <- 1/n # one period
  for (i in seq(n)){
    epsilon <- runif(n=1, min=0, max=1) # random generated number
    # calculate stock price (using quantile function of normal distribution)
    stock_price <- stock_price * (1 + qnorm(epsilon, 
                                            stock_mu * delta_t, 
                                            stock_sigma* sqrt(delta_t)))
  }
  return(stock_price)
}

```

## Run it

```{r, doSimulation}
simulations <- 10000 # number of MC simulations
n <- 20 # trading days
stock_price <- 100
stock_mu <- .1 # drift 10%
stock_sigma <- .2 # volatility 20%
 
# Monte Carlo simulations
set.seed(42) # for reproducibility
stock_prices <- c()
for (i in seq(simulations)){
  stock_prices <- c(stock_prices,
                    f_stock_return(stock_price=stock_price, 
                                   n=n, 
                                   stock_mu=stock_mu, 
                                   stock_sigma=stock_sigma))
}
```

## See the stats

The answer is given by the **quantile** function:  (`r quantile(stock_prices, c(.01, .05))`).

We can see that in 9 900 (99%) scenarios the price is larger than ca. 67 \$; 95% of them were higher than ca. 77 \$. Which gives us needful assessment of how bad our investment could go.

# The approximation of $\pi$

Monte Carlo method can be applied also to problems that can be reformulated to have probabilistic interpretation. Very popular example is the approximation of the number Pi.

This example is based on the fact that if you randomly generate points in a square, π/4 of them should lie within an inscribed circle.

Assuming the perimeter of the circle is r, area of the square is equal to 4r^2^ and area of the inscribed circle is πr^2^. Therefore the probability of a point lying inside the inscribed circle is π/4.

## The simulation

```{r, piSim}

# parameters
simulations <- 10000 # number of simulations
perimeter <- 1
 
# randomly generate a point and check if it is in circle
f_point_in_circle <- function(perimeter=1){
  x <- runif(n=1, min=-perimeter, max=perimeter)
  y <- runif(n=1, min=-perimeter, max=perimeter)
  return(list(x=x, 
              y=y,
              in_circle=x^2 + y^2 <= perimeter^2))
}
 
# Monte Carlo simulations
set.seed(123)
pi_df <- data.frame(x=rep(NA, simulations),
                    y=rep(NA, simulations),
                    in_circle=rep(NA, simulations))
for (i in seq(simulations)){
  my_simulation <- f_point_in_circle()
  pi_df$in_circle[i] <- my_simulation$in_circle
  pi_df$x[i] <- my_simulation$x
  pi_df$y[i] <- my_simulation$y
}
 
my_pi <- 4 * sum(pi_df$in_circle) / nrow(pi_df)

```

And the approximation is: `r sprintf("%.7f",my_pi)` which differs by `r sprintf("%.7f", pi - my_pi)` from R's supplied value.

## Plot it

```{r, plotPi}

library(ggplot2)
ggplot(pi_df, aes(x=x, y=y, color=as.factor(in_circle))) +
  geom_point() +
  theme(legend.position='none') +
  coord_fixed() # fix the aspect ratio
```

[Back to Index](../README.html)