---
title: "Tidy Data"
author: "J. R. Minter"
date: "Started: 2018-06-12, Last modified: 2018-06-13"
output:
  knitr:::html_vignette:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
library(knitr)
# make this an external chunk that can be included in any file
options(width = 72)
# opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'figure/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
suppressWarnings(suppressMessages(suppressPackageStartupMessages(library(tidyverse))))
```
[Back to Index](../README.html)

# Load Packages

```{r loadPackages}
library(tidyverse)
library(here)
```

# Introduction

This data set (`CyclerCPK.csv`) came from Ed Boone's Statistics 321
[YouTube](https://www.youtube.com/watch?v=7j294A4dPlw)
video on how to transform "wide" dataframes to "long" dataframes.
It is available on his
[shared Google Drive](https://drive.google.com/drive/folders/0B5AwHsCad1WCdkRJVHNtX296dW8).

We will use Hadley Wickham's approach to prepare "tidy data"
using `dplyr` (included in) the `tidyverse` meta package.

# Read the data

Read the `CyclerCPK` dataset. Use the `here` package to make path
management easier.

```{r readCsvFile, comment=NA}
fi <- here("tidyData/dat/CyclerCPK.csv")
df <- read.csv(fi, header=TRUE)
head(df)
```

# Tidy the data

Note how the replicate values of `CPK` are listed as columns
`CPK1:CPK3`. We will **rename** those columns to make it easier
to process and then use the function `gather` to make a long dataset
with columns labeled `replicate` and `CPK`.


```{r test}
names(df) <- c("Subject","Age","Gender","TRT",
			   "CPK.1", "CPK.2", "CPK.3", "CPK.4")
dfTidy <-	df %>%
			gather(replicate, CPK, CPK.1:CPK.4)
head(dfTidy)
```

We will now write a function to extract **only the replicate number**.
Let's test with a single value.

```{r, comment=NA}
extractReplicate <- function(x){
	unlist(as.integer(unlist(strsplit(x, "[.]"))[2]))
}
extractReplicate("CPK.1")
```

Now we will use R's `lapply()` function to apply it to the whole
vector of observations. This will return a **list**, that we need
to use `unlist()` to transform into a vector. We will then replace
the `replicate` variable with just the replicate **number** as a
`factor` (categorical variable).

We'll look at the top and bottom of the dataframe to see that we have
a tidy dataframe.

```{r prepForGather, comment=NA}
test <- unlist(lapply(dfTidy$replicate,extractReplicate))
head(test)
dfTidy$replicate = as.factor(test)
head(dfTidy)
tail(dfTidy)
```

And finish with a summary

```{r summary, comment=NA}
summary(dfTidy)
```

[Back to Index](../README.html)
