---
title: "Gnuplot Tips"
author: "J. R. Minter"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  knitr:::html_vignette:
    css: ../theme/jm-blue-vignette.css
    number_sections: yes
    toc: yes
---

[Back to Index](../README.html)

These are John Minter's tips for using gnuplot. They were
compiled from several web sources. Some helpful ones:

[The Gnuplot FAQ](http://www.gnuplot.info/faq/faq.html)

[Gnuplot tricks](http://gnuplot-tricks.blogspot.com/2009/05/gnuplot-tricks-many-say-that-it-is.html)

[gnuplotting.org](http://www.gnuplotting.org/)


# Hacking the .plt file

## Devices

### Use postscript to get both pdf and png output

The best png output comes via postscript. Make sure ImageMagick's
**convert** program is <span style="color:red">**ahead**</span> of windows **convert** program in the
path. Then use this macro to produce <span style="color:red">**both**</span> **pdf** 
<span style="color:red">**and**</span> **png** plots at the same time!

Use something like this:

```
# easy to comment out with #
set terminal postscript enhanced color "Arial" 14
set output "foo.ps"
# proc-gnuplot.sh foo
# set terminal window
```

Where the **proc-gnuplot.sh** script (in C:/Apps/local/path) converts the file to
pdf and crops the margins by using ps2pdf (from texlive) and convert (from ImageMagick) like this:

```
ps2pdf $1.ps temp.pdf
pdfcrop --margins 4 temp.pdf $1.pdf
rm -rf temp.pdf
convert -density 150 -geometry 68.82% $1.pdf $1.png
```
Cropping the margins is especially helpful for inclusion
in $\LaTeX$ documents. This produces a 1024 pixel wide png image of the plot.


One can also this manually using Fiji's "Import PDF" and choosing a 200 dpi from the dialog. This is nice because one can copy an image of a ROI from which an EDS spectrum was collected and paste it into the PNG of the spectrum to make a self-contained figure.


# Useful DTSA Spectrum Plot hacks

```
unset logscale x
set logscale y
set xrange [0.0:10000.0]
set yrange [1.0:5700.0]
set xlabel "Energy [eV]" font "Arial,16"
set ylabel "Intensity" font "Arial,16"
set format x "%g"
set format y "%g"
set key spacing 1.5

unset label
```

Note that one can add a title and split the plot line, like


```

set title "Put key info here..." font "Arial,16"


plot "-" with lines lt 1 lw 2 lc rgb "#9ecae1" \
     title "  0 nm Al on Si 5 kV", \
     "-" with lines lt 1 lw 2 lc rgb "#6baed6" \
     title " 20 nm Al on Si 5 kV", \
     "-" with lines lt 1 lw 2 lc rgb "#4292c6" \
     title " 40 nm Al on Si 5 kV", \
     "-" with lines lt 1 lw 2 lc rgb "#2171b5" \
     title " 60 nm Al on Si 5 kV", \
     "-" with lines lt 1 lw 2 lc rgb "#08519c" \
     title " 80 nm Al on Si 5 kV", \
     "-" with lines lt 1 lw 2 lc rgb "#08306b" \
     title "100 nm Al on Si 5 kV", \
     "-" with impulses lw 2 lt -1 notitle

```

if one only has two elements in a DTSA plt, one can

```
unset label

set label "Sn lines in black, Cl lines in red" at 400,20000

plot "-" with lines lt 1 lw 2 lc rgb "#133463" title "Sn", \
     "-" with lines lt 1 lw 2 lc rgb "#3A89C9" title "SnCl_{2}", \
     "-" with impulses lw 2 lc rgb "red" lt -1 notitle, \
     "-" with impulses lw 2 lc rgb "black" lt -1 notitle
```

followed by (at the end)

```
e
# Cl
2622.4, 34888.0
# Cl
2620.8, 17653.3
# Cl
2815.6, 1604.8
# Cl
2815.6, 3289.9
e
# Sn ***Note the split here...***
619.9, 846.0
# Sn
3444.0, 23114.0
# Sn
577.9, 169.2
# Sn
3435.5, 2621.1
# Sn
3045.0, 961.5
# Sn
4067.5, 62.4
# Sn
3750.3, 3575.7
# Sn
3662.8, 10044.4
# Sn
4132.2, 1340.6
# Sn
690.5, 3384.0
# Sn
490.0, 16.9
# Sn
3971.4, 124.8
# Sn
3272.3, 365.2
# Sn
3708.3, 2191.2
# Sn
396.2, 1692.0
# Sn
732.5, 169.2
# Sn
4463.0, 37.0
# Sn
4376.1, 742.0
# Sn
404.7, 169.2
# Sn
4376.1, 44.6
# Sn
3904.9, 3395.2
# Sn
4019.6, 74.0
e

```



# Color

## RGB Colors

One can use these names
[Hamburg-Schneider](http://www.uni-hamburg.de/Wiss/FB/15/Sustainability/schneider/gnuplot/colors.htm)

## HTML Colors

One can define custom palettes using
[ColorBrewer](http://colorbrewer2.org/)

R makes this easy:

```{r}
library(RColorBrewer)
pal <-brewer.pal(8,"Oranges")
pal
```

## Plotting lines

use these in gnuplot like
```
1 lt rgb "#FFF5EB"
2 lt rgb "#FEE6CE"
3 lt rgb "#FDD0A2"
4 lt rgb "#FDAE6B"
5 lt rgb "#FD8D3C"
6 lt rgb "#F16913"
7 lt rgb "#D94801"
8 lt rgb "#8C2D04"

plot "-" with lines lt 1 lw 2 lc rgb "#FDD0A2" title "100 nm Cu", ...
```

or
```{r}
pal <-brewer.pal(8,"Blues")
pal
```

## Changing label colors

Try

```
set label 3 "Si (ignored)" at first 1730,90000 center textcolor rgb "skyblue"
```

# Adding symbols

```
# Setting the terminal postscript with the options
set terminal postscript eps color enhanced "Helvetica" 20

# Setting the output file name
set output 'plot.eps'

#Setting up the grid and labels
set grid
set title "A simple graphic without LaTeX"
set xrange [-pi:pi]
set xlabel "Angle {/Symbol a}"
set ylabel "{/Symbol Dw}"
set ytics 0.5
set xtics ("{/Symbol p}" -pi, "{/Symbol p}/2" -pi/2, "0" 0, "{/Symbol p}/2" pi/2, "{/Symbol p}" pi)

# Setting the legend
set key horizontal below height 2
set key box lt 2 lc -1 lw 3

# The plot itself (\ is to broke lines)
set size 1,1
plot sin(x)**2 w l lt 1 lc 1 lw 3 t "sin^2({/Symbol a})", \
sin(2*x)/x w l lt 2 lc 3 lw 3 t "sin(2{/Symbol a})/{/Symbol a}", \
sin(x) w l lt 3 lc 2 lw 3 t "sin{/Symbol a})"

```

# Controlling borders

The FAQ suggests:

```
set lmargin at screen 0.05
set bmargin at screen 0.05
set rmargin at screen 0.95
set tmargin at screen 0.95
```



[Back to Index](../README.html)