Gnuplot tips
========================================================

[Back to Index](../README.html)

These are John Minter's tips for using gnuplot. They were
compiled from several web sources. Some helpful ones:

[The Gnuplot FAQ](http://www.gnuplot.info/faq/faq.html)

[Gnuplot tricks](http://gnuplot-tricks.blogspot.com/2009/05/gnuplot-tricks-many-say-that-it-is.html)


# Hacking the .plt file

## Devices

### Postscript

Use something like this
```
# easy to comment out with #
# set terminal postscript enhanced "Arial" 16
# set output "200nmNiOnCu15kV.ps"
# set terminal png
# set terminal window


```

## Useful DTSA Spectrum Plot hacks

```
unset logscale x
set logscale y
set xrange [0.0:10000.0]
set yrange [1.0:5700.0]
set xlabel "Energy [eV]" font "Arial,16"
set ylabel "Intensity" font "Arial,16"
set format x "%g"
set format y "%g"
unset label
```

## HTML Colors

One can define custom palettes using
[ColorBrewer](http://colorbrewer2.org/)

R makes this easy:

```{r}
library(RColorBrewer)
pal <-brewer.pal(8,"Oranges")
pal
```

use these in gnuplot like
```
1 lt rgb "#FFF5EB"
2 lt rgb "#FEE6CE"
3 lt rgb "#FDD0A2"
4 lt rgb "#FDAE6B"
5 lt rgb "#FD8D3C"
6 lt rgb "#F16913"
7 lt rgb "#D94801"
8 lt rgb "#8C2D04"
```

or
```{r}
pal <-brewer.pal(8,"Blues")
pal
```
### Controlling borders

The FAQ suggests:

```
set lmargin at screen 0.05
set bmargin at screen 0.05
set rmargin at screen 0.95
set tmargin at screen 0.95
```

### Adding symbols

```
# Setting the terminal postscript with the options
set terminal postscript eps color enhanced "Helvetica" 20

# Setting the output file name
set output 'plot.eps'

#Setting up the grid and labels
set grid
set title "A simple graphic without LaTeX"
set xrange [-pi:pi]
set xlabel "Angle {/Symbol a}"
set ylabel "{/Symbol Dw}"
set ytics 0.5
set xtics ("{/Symbol p}" -pi, "{/Symbol p}/2" -pi/2, "0" 0, "{/Symbol p}/2" pi/2, "{/Symbol p}" pi)

# Setting the legend
set key horizontal below height 2
set key box lt 2 lc -1 lw 3

# The plot itself (\ is to broke lines)
set size 1,1
plot sin(x)**2 w l lt 1 lc 1 lw 3 t "sin^2({/Symbol a})", \
sin(2*x)/x w l lt 2 lc 3 lw 3 t "sin(2{/Symbol a})/{/Symbol a}", \
sin(x) w l lt 3 lc 2 lw 3 t "sin{/Symbol a})"

```

[Back to Index](../README.html)