---
title: "Gnuplot Tips"
author: "J. R. Minter"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  knitr:::html_vignette:
    css: ../theme/jm-mod-vignette.css
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Gnuplot Tips}
  %\VignetteEngine{knitr::rmarkdown}
---

[Back to Index](../README.html)

These are John Minter's tips for using gnuplot. They were
compiled from several web sources. Some helpful ones:

[The Gnuplot FAQ](http://www.gnuplot.info/faq/faq.html)

[Gnuplot tricks](http://gnuplot-tricks.blogspot.com/2009/05/gnuplot-tricks-many-say-that-it-is.html)

[gnuplotting.org](http://www.gnuplotting.org/)


# Hacking the .plt file

## Devices

### Postscript

Use something like this
```
# easy to comment out with #
# set terminal postscript enhanced color "Arial" 16
# set output "200nmNiOnCu15kV.ps"
# proc-gnuplot.sh 200nmNiOnCu15kV
# set terminal png
# set terminal window
```

Note the proc-gnuplot.sh script converts the file to pdf and crops the 
margins like this:

```
ps2pdf 200nmNiOnCu15kV.ps temp.pdf
pdfcrop --margins 4 temp.pdf 200nmNiOnCu15kV.pdf
rm -rf temp.pdf
```
Cropping the margins is especailly helpful for inclusion
in $\LaTeX$ documents.

### PNG
The best png output comes via postscript. Make sure ImageMagick's
``convert`` program is ahead of windows ``convert`` program in the
path. Then use

```
convert -density 150 -geometry 100% plt.ps plt.png
```

One can so this manually using Fiji's "Import PDF" and chosing a 200 dpi from the dialog. This is nice because one can copy an image of a ROI from which an EDS spectrum was collected and paste it into the PNG of the spectrum to make a self-contained figure.


# Useful DTSA Spectrum Plot hacks

```
unset logscale x
set logscale y
set xrange [0.0:10000.0]
set yrange [1.0:5700.0]
set xlabel "Energy [eV]" font "Arial,16"
set ylabel "Intensity" font "Arial,16"
set format x "%g"
set format y "%g"
unset label
```

# Color

## RGB Colors

One can use these names
[Hamburg-Schneider](http://www.uni-hamburg.de/Wiss/FB/15/Sustainability/schneider/gnuplot/colors.htm)

## HTML Colors

One can define custom palettes using
[ColorBrewer](http://colorbrewer2.org/)

R makes this easy:

```{r}
library(RColorBrewer)
pal <-brewer.pal(8,"Oranges")
pal
```

### Plotting lines

use these in gnuplot like
```
1 lt rgb "#FFF5EB"
2 lt rgb "#FEE6CE"
3 lt rgb "#FDD0A2"
4 lt rgb "#FDAE6B"
5 lt rgb "#FD8D3C"
6 lt rgb "#F16913"
7 lt rgb "#D94801"
8 lt rgb "#8C2D04"

plot "-" with lines lt 1 lw 2 lc rgb "#FDD0A2" title "100 nm Cu", ...
```

or
```{r}
pal <-brewer.pal(8,"Blues")
pal
```

## Changing label colors

Try

```
set label 3 "Si (ignored)" at first 1730,90000 center textcolor rgb "skyblue"
```

# Adding symbols

```
# Setting the terminal postscript with the options
set terminal postscript eps color enhanced "Helvetica" 20

# Setting the output file name
set output 'plot.eps'

#Setting up the grid and labels
set grid
set title "A simple graphic without LaTeX"
set xrange [-pi:pi]
set xlabel "Angle {/Symbol a}"
set ylabel "{/Symbol Dw}"
set ytics 0.5
set xtics ("{/Symbol p}" -pi, "{/Symbol p}/2" -pi/2, "0" 0, "{/Symbol p}/2" pi/2, "{/Symbol p}" pi)

# Setting the legend
set key horizontal below height 2
set key box lt 2 lc -1 lw 3

# The plot itself (\ is to broke lines)
set size 1,1
plot sin(x)**2 w l lt 1 lc 1 lw 3 t "sin^2({/Symbol a})", \
sin(2*x)/x w l lt 2 lc 3 lw 3 t "sin(2{/Symbol a})/{/Symbol a}", \
sin(x) w l lt 3 lc 2 lw 3 t "sin{/Symbol a})"

```

# Controlling borders

The FAQ suggests:

```
set lmargin at screen 0.05
set bmargin at screen 0.05
set rmargin at screen 0.95
set tmargin at screen 0.95
```



[Back to Index](../README.html)