---
title: "R-caracas"
author: "J. R. Minter"
date: "Started: 2020-05-26, Last modified: 2020-05-26"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, comment=NA}
library(knitr)
options(width = 72)
suppressPackageStartupMessages(library("tidyverse"))
```


[Back to Index](../README.html)

`caracas` is an R package by Mikkel Anderson and Soren Hojsgaard.
`caracas` implements computer algebra built on `SymPy` using
`reticulate`. `caracas` is on CRAN.

I found out about caracas on
[Twitter](https://twitter.com/mikldk/status/1265168063823073282)

Make sure `SymPy` is installed by

```
conda install SymPy
```

Note: I specified the python interpreter in my `.Renvion` file
with:

```
RETICULATE_PYTHON="/Users/jrminter/miniconda3/bin"
```

because my MacBook Pro has both system python and minconda python... 

# Example: Solve an equation

```{r caracasExample1, comment=NA}
library(reticulate)
reticulate::py_config()
```
```{r testOne, message=FALSE, comment=NA}

library(caracas)

x <- symbol("x")
y <- symbol("y")
solve_sys(x^2 - y, x)
```

# Example: Multinomial likelihood

```{r caracasExample2, comment = NA}
p <- as_symbol(paste0("p", 1:3))
y <- as_symbol(paste0("y", 1:3))
a <- as_symbol("a")
l <- sum(y*log(p))
L <- -l + a*(sum(p) - 1)
gL <- der(L, c(a, p))
sol <- solve_sys(gL, c(a,p))
sol
```

```{r caracasExample3, comment=NA}
A <- as_symbol(diag(3))
diag(A)[-2] <- "b"
A[2, 1] <- "b"
A

inv(A) %*% A

tex(A)
```


[Back to Index](../README.html)
