---
title: "Bar Plots in R"
author: "J. R. Minter"
date: "Started: 2013-10-21, Last modified: 2020-04-12"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

[Back to Index](../README.html)

# Bar Chart Examples by Bruno Rodrigues

Bruno reported these examples on 2020-04-12

> Bar charts are quite simple plots, but there are enough variations of them
> that they deserve one single blog post. However, don’t expect many
> explanations.

> Let’s first start by loading some data, and the usually required packages:

```{r loadRequiredPackages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(colorspace)
library(ggplot2)
```

## Load a data set

```{r loadData}
data(gss_cat)
```

> Very often, what one wants to show are counts:

```{r getCounts}
gss_cat %>%
  count(marital, race)
```

## Make an `other` category

> Let’s lump marital statuses that appear less than 10% of the time into an
> “Other” category:

```{r lumpCounts}
(
  counts_marital_race <- gss_cat %>%
    mutate(marital = fct_lump(marital, prop = 0.1)) %>%
    count(marital, race)
)
```

## Make the simplest bar plot

**_Nota Bene_**: In these examples I have replaced `brotools::theme_blog()`
with `theme_minimal()`.

```{r simplestBarPlot}

simple_plt <- ggplot(counts_marital_race) +
              geom_col(aes(x = marital, y = n, fill = race)) +
              scale_fill_discrete_qualitative(palette = "Dark 3") +
              theme_minimal() +
              NULL

print(simple_plt)
```

## Now with position = "dodge":

```{r simpleBarPlot2}
smp_plt_2 <- ggplot(counts_marital_race) +
             geom_col(aes(x = marital, y = n, fill = race),
                      position = "dodge") +
             scale_fill_discrete_qualitative(palette = "Dark 3") +
             theme_minimal() +
             NULL

print(smp_plt_2)
```

## Moving the legend around

We will use `theme(legend.position = ...)`:

```{r themeLegendPosition}
smp_plt_3 <- ggplot(counts_marital_race) +
             geom_col(aes(x = marital, y = n, fill = race),
                      position = "dodge") +
             scale_fill_discrete_qualitative(palette = "Dark 3") +
             theme_minimal() +
             theme(legend.position = "left") +
             NULL

print(smp_plt_3)

```

## Counting by year

We can count by year as well:

```{r countByYear}
(
  counts_marital_race_year <- gss_cat %>%
    mutate(marital = fct_lump(marital, prop = 0.1)) %>%
    count(year, marital, race) %>%
    ungroup()
)

```

## Track a variable change with time

> When you want to show how a variable evolves through time,
> an **area chart** is handy:

```{r AreaChart}

smpl_plt_4 <- counts_marital_race_year %>%
              group_by(year, marital) %>%
              summarise(n = sum(n)) %>%
              ggplot() +
              geom_area(aes(x = year, y = n, fill = marital)) +
              scale_fill_discrete_qualitative(palette = "Dark 3") +
              theme_minimal() +
              NULL

print(smpl_plt_4)

```

## Now use `facets`

```{r useFacets}
smpl_plt_5 <- counts_marital_race_year %>%
              ggplot() +
              geom_area(aes(x = year, y = n, fill = marital)) +
              facet_wrap(facets = vars(race), ncol = 1) +
              scale_fill_discrete_qualitative(palette = "Dark 3") +
              theme_minimal() +
              NULL

print(smpl_plt_5)

```

## `Facets` with individual `y-axis` scales

```{r indAxisYscales}
smpl_plt_6 <- counts_marital_race_year %>%
              ggplot() +
              geom_area(aes(x = year, y = n, fill = marital)) +
              facet_wrap(facets = vars(race), ncol = 1, scales = "free_y") +
              scale_fill_discrete_qualitative(palette = "Dark 3") +
              theme_minimal() +
              NULL

print(smpl_plt_6)

```

## An area chart but with **relative frequencies**

```{r areaChartRelFreq}

smpl_plt_7 <- counts_marital_race_year %>% 
              group_by(year, marital) %>% 
              summarise(n = sum(n)) %>%  
              mutate(freq = n/sum(n)) %>% 
              ggplot() +
              geom_area(aes(x = year, y = freq, fill = marital)) +
              scale_fill_discrete_qualitative(palette = "Dark 3") +
              theme_minimal() +
              NULL

print(smpl_plt_7)
```

## With `facet_wrap()`

```{r withFacetWrap}
smpl_plt_8 <- counts_marital_race_year %>% 
              group_by(race, year, marital) %>% 
              summarise(n = sum(n)) %>%  
              mutate(freq = n/sum(n)) %>% 
              ggplot() +
              geom_area(aes(x = year, y = freq, fill = marital)) +
              facet_wrap(facets = vars(race), ncol = 1, scales = "free_y") +
              scale_fill_discrete_qualitative(palette = "Dark 3") +
              theme_minimal() +
              NULL

print(smpl_plt_8)

```

## Fixing dates...

> Want to replace 2000 with “2000-01-01”? First need to create vector of
> prettier dates and positions?

### Create better dates

```{r createBetterDates}
pretty_dates <- counts_marital_race_year %>%
                mutate(pretty_dates = paste0(year, "-01-01")) %>%
                pull(pretty_dates) %>%
                unique()

position_dates <- counts_marital_race_year %>%
                  pull(year) %>%
                  unique() %>%
                  sort()
```

`scale_x_continuous()` can now use this. Using `guide = guide_axis(n.dodge = 2)`
to avoid overlapping labels:

```{r noOverLapLabels}
smpl_plt_9 <- counts_marital_race_year %>% 
              group_by(race, year, marital) %>% 
              summarise(n = sum(n)) %>%  
              mutate(freq = n/sum(n)) %>%
              ggplot() +
              geom_area(aes(x = year, y = freq, fill = marital)) +
              facet_wrap(facets = vars(race), ncol = 1, scales = "free_y") +
              scale_x_continuous("Year of survey", labels = pretty_dates,
                                  breaks = position_dates,
                                  guide = guide_axis(n.dodge = 2)) +
              scale_fill_discrete_qualitative(palette = "Dark 3") +
              theme_minimal() +
              NULL
print(smpl_plt_9)
```

### Getting labels right is a **major pain** in this case.

The author needed to do some rearanging (see the original...)

```{r rearrangedLabels}

smpl_plt_10 <- counts_marital_race_year %>% 
               group_by(race, year, marital) %>% 
               summarise(n = sum(n)) %>%  
               mutate(freq = n/sum(n)) %>%
               group_by(year, race) %>%  
               mutate(marital = fct_reorder(marital, freq)) %>% 
               arrange(desc(marital)) %>% 
               mutate(position = cumsum(freq)) %>%
               mutate(prev_pos = lag(position, default = 0)) %>%
               mutate(position = (position + prev_pos)/2) %>%  
               ggplot() +
               geom_area(aes(x = year, y = freq, fill = marital)) +
               facet_wrap(facets = vars(race), ncol = 1, scales = "free") +
               scale_x_continuous("Year of survey", labels = pretty_dates,
                                   breaks = position_dates,
                                   guide = guide_axis(n.dodge = 2)) +
               geom_label(aes(x = year, y = position,
                              label = round(100 * freq))) +
               scale_fill_discrete_qualitative(palette = "Dark 3") +
               theme_minimal() +
               NULL

print(smpl_plt_10)
```

I think that is getting pretty complicated...

## Consider the `tv_hours` varible

Now let’s focus on the variable tvhours. We want to show the total watched
hours, but also the total across all the categories of race and marital in a
faceted bar plot:

```{r tvHours}
(
  total_tv <- gss_cat %>%
    group_by(year, race, marital) %>%
    summarise(total_tv = sum(tvhours, na.rm = TRUE))
)
```

This tibble has the total watched hours by year, race and marital status
variables. How to add the total by year and race categories? For this, by are
first going to use the `group_split()`:


**NOTE** I could not get this part to work on R-3.6.3 on mac. I'm tired...

```
{r tvGroupSplit, warning=FALSE}

total_tv_split <- total_tv %>%
                  select(race, year, marital, total_tv) %>%
                  mutate(year = as.character(year)) # %>%  
                  # group_split(year, race)

print(total_tv_split)
```


I have to re-order the columns with select(), because when using
janitor::adorn_totals(), which I will be using below to add totals, the
first column must be a character column (it serves as an identifier column).

This creates a list with 3 races times 6 years, so 24 elements. Each element of 
the list is a tibble with each unique combination of year and race:



[Back to Index](../README.html)
