---
title: "Sweave Tips"
author: "J. R. Minter"
date: "Started: 2013-07-16, Last modified: 2018-10-11"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r, echo=FALSE}
library(knitr)
```

[Back to Index](../README.html)

These are John Minter's tips for a workflow using R/Sweave/LaTeX.
They were compiled from several web sources.

# xtable tips

[StackOverflow](http://stackoverflow.com/questions/19846796/adding-titles-to-xtable) has a nice example.

```
library(xtable)
Grade3 <- c("A","B","B","A","B","C","C","D","A","B","C","C","C","D","B","B","D","C","C","D")
Grade6 <- c("A","A","A","B","B","B","B","B","C","C","A","C","C","C","D","D","D","D","D","D")
Cohort <- table(Grade3,Grade6)
addtorow <- list()
addtorow$pos <- list()
addtorow$pos[[1]] <- 0
addtorow$pos[[2]] <- 0
addtorow$command <- c('& \\multicolumn{4}{c}{Grade 6} \\\\\n', "Grade 3 & A & B & C & D \\\\\n")
print(xtable(Cohort, caption = 'My Title'), caption.placement = 'top', add.to.row = addtorow, include.colnames = FALSE)
```

A simple wrapper that works... Note the renaming of columns
and specifying digits. Remember there that there need to
be ``ncol+1`` entries in the digit vector. Still need to
learn how to put a ``%`` in a label

```
<<echo=FALSE, results=tex>>=
o.col.names <- colnames(my.df)
colnames(my.df) <- c("sample",
                        "plate (s)",
                        "pass (s)",
                        "inc 85",
                        "mean",
                        "std err",
                        "lcl",
                        "ucl")
dig <- c(0,0,0,0,0,2,2,2,2)
print(xtable(my.df, digits = dig,
             caption = "My Caption.",
             label = "tab:mine",
             table.placement = "tbp",
             caption.placement = "top",
             size = "small",
             table.placement="ht"), include.rownames=FALSE)
colnames(my.df) <- o.col.names
@
```

Note: one can avoid Sweave and write the tables as ``.tex`` files
directly from an R script...

```{r}
library(xtable)
df <- data.frame(one=c(1,2,3),
                 two=c(1,2,3),
                 three=c(1,2,3),
                 four=c(1,2,3))

tab.tex  <- './tab.tex'
o.names <- colnames(df)
# if one wants clean column names...
n.names <- c("uno", "dos", "tres", "quatro")
colnames(df) <- n.names
# note ncol+1 entries
dig <- c(0,1,1,1,1)
xt <- xtable(df, digits = dig,
             caption = "My table.",
             label = "tab:myTab",
             table.placement = "tbp",
             caption.placement = "top",
             size = "small",
             table.placement="ht")

sink(tab.tex)
# change rownames to TRUE if meaningful
print(xt, include.rownames=FALSE)
sink()

# the lets one include the file...
str.ei <- '\\endinput'
cat(str.ei, file=tab.tex, sep='\n', append=TRUE)


```


[Back to Index](../README.html)


