---
title: "R data cleaning"
author: "J. R. Minter"
date: "Started: 2019-04-30, Last modified: 2020-04-16"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    csl: ../inc/american-chemical-society-with-titles-sentence-case-doi.csl
    number_sections: yes
    toc: yes
    toc_depth: 3
bibliography: ../inc/tips.bib
---

[Back to Index](../README.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

# Introduction

We get data from many sources. Often the the data needs to be imported from
files where the names and or formats do not play well with R [@r-project].
Hadley Wickham has created the concept of **Tidy Data** [@tidy-data]. The
process we use is called **data cleaning**. There is an excellent R package,
`janitor` by Sam Firke [@janitor] that helps automate the process.

Start by loading the packages we will use...

```{r load_1, message=FALSE}
library(dplyr)
library(janitor)
library(here) # make paths easier...
library(ggplot2) # for plotting
```


# Example 1

Let's consider a simple example from the package vignette. We start by creating
a data frame with "**dirty names**"... Note the spaces in the names and unicode
characters. R does not like those...

```{r, df_w_dirty_names}
test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")
test_df
```

Clean the variable names, returning a data.frame:

```{r clean_the_names}
test_df %>%
  clean_names()
```

Note: I have a problem with the develk version of dplyr on Windows...


# An example from a Casino simulation of "guilding metal"

`Guilding metal` is an alloy of Cu and Zn. I used the Casino v. 2.5.1 software
[the latest major version of [@casino-v2.42]] to model the X-ray emission from
the alloy when exposed to a 25 kV electron probe.

The result are two tab-delimited files, `casino-guilding-Zn-25kV.dat` and
`casino-guilding-Cu-25kV.dat`. We need to skip the first 4 lines that have
a title (prz of Cu or Zn), and then our five column names. Let's read in the
Cu data

## Process the Cu data

```{r read_cu_data}

path <- here("./R-data-cleaning/dat/casino-guilding-Cu-25kV.dat")

df <- read.table(path, header = TRUE, sep="\t", skip=4)
print(head(df))

clean_cu <- df %>%
            clean_names()
head(clean_cu)
```

## Process the Zn data

Now, we will repeat the process for the Zn data


```{r read_zn_data}

path <- here("./R-data-cleaning/dat/casino-guilding-Zn-25kV.dat")

df <- read.table(path, header = TRUE, sep="\t", skip=4)
clean_zn <- df %>%
            clean_names()
head(clean_zn)
```

## Plot the $\phi (\rho z)$ curve

```{r plot_k_data}

plt <-  ggplot() +
        geom_point(data=clean_cu, aes(x=depth_nm, y=intensity_k),
                   colour="darkblue") +
        annotate("text", label = 'Cu K-L3', x = 1000., y = 1.75,
                 size = 5, colour = "darkblue") +
        geom_point(data=clean_zn, aes(x=depth_nm, y=intensity_k),
                  colour='darkred') +
        annotate("text", label = 'Zn K-L3', x = 1000., y = 1.25,
                 size = 5, colour = "darkred") +
        xlab(label="depth [nm]") +
        ylab(label="\U03D5(\U03C1z)") +
        ggtitle("Casino \U03D5(\U03C1z) Model of Guilding Metal at 25 kV") +
        theme(axis.text=element_text(size=12),
              axis.title=element_text(size=14),
              plot.title = element_text(hjust = 0.5)) + # center the title
        
        NULL
plt
```





[Back to Index](../README.html)

# References