---
title: "R Matrix Algebra"
author: "J. R. Minter"
date: "Started: 2018-06-11, Last modified: 2018-06-11"
output:
  knitr:::html_vignette:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
---

[Back to Index](../README.html)

# Introduction

R has operators and functions for efficient matrix algebra.

# Vectors

R lets us define vectors

```{r DefineTwoVectors}
a1 <- c(1,3,4,2,5)
b1 <- c(2,1,3,5,4)
```

## Simple addition of vectors

We can add a **constant to each element.**


```{r, comment=NA}
a5 <- a1 + 5
print(a1)
print(a5)
```

Note that if we add `a1` and `b1` together, they add
**element by element**!


```{r, comment=NA}
ab <- a1 + b1
print(ab)
```

## Simple multiplication of vectors

```{r, comment=NA}
print(a1)
print(5*a1)  # multiplies by a scalar
a_times_b <- a1*b1
print(a_times_b) # element by element multiplication
```

Note how we select (and print) a **single element** of a vector.

```{r, comment=NA}
print(a_times_b[2]) # print the second element
```

# Matrices

Note that R fills the matrix **column-by-column**.

```{r, comment=NA}
m1 <- matrix(seq(1:4), nrow=2, ncol=2)
m2 <- matrix(c(4,3,2,1), nrow=2, ncol=2)
print(m1)
```

## True matrix multiplication

For true matrix multiplication, you need an `m x n` matrix times an
`n x p` matrix.

```{r, comment=NA}
n1 <- matrix(seq(1:8), nrow=2, ncol=4)
n2 <- matrix(seq(1:24), nrow=4, ncol=6)
n1
n2
n1 %*% n2
```

[Back to Index](../README.html)