---
title: "R memes"
author: "J. R. Minter"
date: "Started: 2019-04-17, Last modified: 2019-04-17"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, include = T, cache = F, echo = F, message = F, warning = F, tidy = F}
library(knitr)
library(meme)
library(grid)
library(ggplot2)

options(width = 72)

# suppressPackageStartupMessages(library("tidyverse"))
```


[Back to Index](../README.html)

# Introduction

Like it or not, memes are a part of communication on social media. Happily,
there is a R package (`meme`) that helps.

# Getting started

[Guangchuang Yu](https://cran.r-project.org/web/packages/meme/vignettes/meme.html#special-note-for-windows-users)
wrote a nice introduction to help us get started.

Windows users need to register fonts in a chunk like this

```{r windows-set_font}
if (.Platform$OS.type == "windows") {
    windowsFonts(
        Impact = windowsFont("Impact"),
        Courier = windowsFont("Courier")
    )
}
```

# meme

Call meme to add meme captions like this:

```{r make_angry_meme}
u <- system.file("angry8.jpg", package="meme")
meme(u, "code", "all the things!")
```

# The grammar of meme

Not that useful, just to mimic ggplot2:

```{r mimic_ggplot2}
mmplot(u) + mm_caption("calm down", "and RTFM",
                       color="purple")
```

# meme_save: a meme version of ggsave

The meme output can be saved as an object, and can be exported to file using
meme_save. meme_save helps user setting up the output figure aspect ratio and
calls ggsave to export the figure:

```{r meme_save_example}
u2 <- system.file("success.jpg", package="meme")
x <- meme(u2, "please", "tell me more")
outfile <- tempfile(fileext=".png")
meme_save(x, file=outfile)
```

# `plot` method

Users can plot the meme output and change the caption or other parameters in
real time.

```{r plot_method}
plot(x, size = 2, "happy friday!", "wait, sorry, it's monday", color = "firebrick", font = "Courier")
```

# `+` method

Instead of using parameters in `plot()` explictely, Users can use `+ aes()`
to set the plot parameters:

```{r use_aes}
x + aes(upper = "#barbarplots",
        lower = "friends don't let friends make bar plots",
        color = firebrick, font = Courier, size=1.5)
```


or using `+ list().` The following command will also generate the figure
displayed above.

```{r plus_list}
x + list(upper = "#barbarplots",
        lower = "No bar plots, please!",
        color = "firebrick", font = "Courier", size=1.5)
```

# multi-language support

I didn’t do anything about it. Multi-language was supported internally. Just simply select a font for your language.

I do not have a suitable font...

```
{r multi_language}
y <- meme(u, "卧槽", "听说你想用中文", font="STHeiti")
y
```

# grid support

```{r need_for_grid_and_other}

xx <- seq(0, 2*pi , length.out=10)
yy <- sin(xx)
mm <- meme(u, "code", "all the things!", size=.3, color='firebrick', bgcolor=NULL)
```


This does not work well

```
{r grid_support}
grid.newpage()
pushViewport(viewport(width=.9, height=.9))
grid.rect(gp = gpar(lty="dashed"))


for (i in seq_along(xx)) {
    vp <- viewport(x = xx[i]/(2*pi), y = (yy[i]-min(yy))/2, width=.05, height=.05)
    print(mm, vp = vp)
}

```


# ggplot2 support

```{r ggplot2_support}

library(ggplot2)
library(ggimage)

d <- data.frame(x = xx, y = yy)
ggplot(d, aes(x, y)) + geom_line() +
    geom_subview(aes(x, y), data=d, subview=mm, width=.3, height=.15)

ggplot(d, aes(x, y)) +
    geom_subview(x = 0, y = 0, subview=mm+aes(size=3), width=Inf, height=Inf) +
    geom_point() + geom_line()

```

# cowplot support

This does not run...
```
{r cow_plot_support}
library(cowplot)
plot_grid(x, y, ncol=1, labels = c("A", "B"))

```

# font support

In addition to the fonts installed on your system, there are many amusing fonts that are funny for meme, e.g. fonts on https://fontmeme.com/fonts/.

```{r import_pokemon_fonts}
## import pokemon fonts
## downloaded from <https://fontmeme.com/fonts/pokmon-font/>
font_pokemon()
## font Pokemon_Hollow, Pokemon_Solid loaded...
u <- 'https://ravishly.com/sites/default/files/landscape-1456483171-pokemon2.jpg'
meme(u, "Pokemon", "pikachu i choose you!", font='Pokemon_Hollow')

meme(u, "Pokemon", "pikachu i choose you!", font='Pokemon_Solid', color='#FCCF00')
```

You can download fonts and pass the folder to font_import() to import the fonts\ so that they can be used to create meme.

```{r download_bubble}
## folder that contains bubble1 font
## downloaded from https://fontmeme.com/fonts/bubble-1-font/
dir <- system.file('fonts/bubble', package='meme')
font_import(dir)
## font bubble1 loaded...
meme(u, "the meme package", "is awesome!", font="bubble1")
```

Fonts imported by `font_import()` are not restricted in creating meme,
you can use them in other visualization packages.

**Note:** When I included a line like:

```
  theme(plot.title=element_text(family='bubble1', size=30, color='firebrick'))
```
I got an error.

This worked

```{r reuse_fonts, error=FALSE}
qplot(1:10, 1:10) +
  labs(title="meme is awesome") +
  theme(plot.title=element_text(size=30,
                                color='firebrick'))
```

# Session Info

```{r session_info}
sessionInfo()
```




[Back to Index](../README.html)
