---
title: "R model prediction tips"
author: "J. R. Minter"
date: "Started: 2018-03-15, Last modified: 2018-03-15"
output:
  knitr:::html_vignette:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
---

[Back to Index](../README.html)

# Overview

I frequently need use R's model fitting capabilities and then want to
compute an interpolation for a smooth curve. This document contains
some reproducible examples.

# Fitting Misell and Burdett's data

To prepare the data for modeling, we need to load and pre-process a
dataset from Misell and Burdett. We need to do the following
transformations:

1. Convert $kV$ to $eV$ (using `fac1` below)
2. Convert $m^2/mg$ to $cm^2/g$ (using `fac2` below)
3. Take the natural log of each term. *N.B.* In R the natural log
function is `log()` while the base 10 function is `log10()`.

```{r loadAndPrepData, message=FALSE}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(pander))
suppressPackageStartupMessages(library(tidyverse))
fi <- here('R-model-predict/dat/Misell-Burdett.csv')
df <- read_csv(fi)
fac1 <- 1000 # kev to eV
fac2 <- 10000000 # m2/mg to cm2/g 
df.ln <- data.frame(ln.E0.eV=log(fac1*df[,1]),
                    ln.alpha.05=log(fac2*df[,2]),
                    ln.alpha.10=log(fac2*df[,3]),
                    ln.alpha.15=log(fac2*df[,4]))
names(df.ln) <- c("ln.e0.eV","ln.alpha.05",
                  "ln.alpha.10","ln.alpha.15")
pander(df.ln)
```

Next, let's fit the data to a linear model for the 5 mrad aperture.
Note how we create a dataframe for the model and do the fit. We get
the coefficients from the `summary()` function.

```{r, fitAp05}
df.05 <- data.frame(x=df.ln[,1], y=df.ln[,2])
lm.05 <- lm(y~x, data=df.05)
sum.05 <- summary(lm.05)
coe.05 <- sum.05$coef
pander(coe.05)
```

We now define a function to get the slope and intercept from the
coefficients so we can tabulate them. We would re-use these functions
to extract coefficients from fits to other apertures with different
scattering angles.

```{r, getCoef05}
getInterceptFromCoef <- function(coe){
  mu <- coe[1]
  se <- coe[3]
  ret <- c(mu,se)
  names(ret) <- c("mu", "se")
  return(ret)
}

getSlopeFromCoef <- function(coe){
  mu <- coe[2]
  se <- coe[4]
  ret <- c(mu,se)
  names(ret) <- c("mu", "se")
  return(ret)
}

sl.05 <- getInterceptFromCoef(coe.05)
in.05 <- getSlopeFromCoef(coe.05)
```
And compute the predicted values for the line.
**Note how we define a vector of x-values**
using the `seq()` function and insert it into our
working dataframe (`df.05.p`). We then fill in the predicted values
with the `predict()` function. The key is to pass the dataframe name
to the function using the `newdat` parameter.

```{r, predAp05}
x.pred <- seq(from=9.90, to=11.6, by=0.01)
df.05.p <- data.frame(x=x.pred)
df.05.p$y <- predict(lm.05, newdat=df.05.p)
```

Next, we plot the data.

```{r, makeLogPlot}

plt <- ggplot() +
       geom_point(data=df.ln, aes(x=ln.e0.eV, y=ln.alpha.05),
                  colour="darkblue") +
       geom_line(data=df.05.p, aes(x=x, y=y), colour="darkblue") +
       xlab(label="ln(e0 [eV])") +
       ylab(label="ln(Sp [cm2/g])") +
       ggtitle("Misell-Burdett data") +
       theme(axis.text=element_text(size=12),
             axis.title=element_text(size=12),
             # center the title
             plot.title = element_text(hjust = 0.5))

print(plt)
```

# Prediction using LOESS

Let's first illustrate the key methodology from a data set of the Al-K
and Si-K X-ray peak intensities computed by Monte Carlo simulation of
layers of selected thicknesses of Al on a Si substrate using the NIST
DTSA-II X-ray EDS software. We want to use this to estimate the
thickness of vacuum-deposited films of Al by attaching a small chip of
Si to the surface of the coated material and then recording X-ray EDS
spectra at 5 kV.

Let's start by loading a data-frame from disk and compute the derived
quantity we need to plot. We want to use the LOESS algorithm to compute
a smooth, monotonic curve. The values from this curve may be used as a
look-up table to compute thickness from measured spectra.

```{r, message=FALSE}
fi <- here('R-loess/dat/Al-on-Si-5kV-10000-Traj.csv')
df <- read_csv(fi)
df$al.to.si.mu <- df$Al.Int.mu / df$Si.Int.mu
df$al.to.si.unc <- sqrt((df$Al.Int.unc/df$Al.Int.mu)**2 + 
                        (df$Si.Int.unc/df$Si.Int.mu)**2)
```

Now compute the LOESS, using the default span.

```{r, computeLoess}
y.loess <- loess(al.to.si.mu ~ t.Al.nm, span=0.75, data=df)
```

Next, we want to **predict the smooth curve with high resolution**.

```{r, predictLoess}
t.Al.nm <- seq(from=0.5, to=100, by=0.5)
df2 <- data.frame(t.Al.nm=t.Al.nm)
df2$al.to.si.mu = round(predict(y.loess, newdata = df2), 5)
```

Next prepare a ggplot2 plot, named `altosiInt`:

```{r}
altosiInt <- ggplot(df, aes(x=t.Al.nm, y=al.to.si.mu)) + 
    geom_errorbar(aes(ymin=al.to.si.mu - 1.96*al.to.si.unc,
                      ymax=al.to.si.mu + 1.96*al.to.si.unc), width=.1) +
    geom_line(color='red', size=1.25, aes(color="red"), data=df2) +
    annotate("text", label = 'loess',
             x = 80, y = 2.25,
             size = 5, colour = "red") +
    geom_point(color='black', size=2) +
    xlab("Al thickness [nm]") +
    ylab("Al-K/Si-K X-ray intensity ratio") +
    ggtitle("Monte Carlo Model of Al on Si at 5 kV") +
    theme(axis.text=element_text(size=12),
          axis.title=element_text(size=14),
    	  plot.title = element_text(hjust = 0.5))# or ,face="bold")) +

print(altosiInt)
```


[Back to Index](../README.html)