---
title: "Tips on Using Color"
author: "J. R. Minter"
date: "Started: 2013-07-23, Last modified: 2018-07-15"
output:
  knitr:::html_vignette:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
---
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
library(knitr)
options(width = 72)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'figure/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


[Back to Index](../README.html)


Color palettes need to be chosen carefully.

# Resources

- **Creating corporate color palettes for ggplot2** by [\@drsimonj](https://drsimonj.svbtle.com/creating-corporate-colour-palettes-for-ggplot2)
Simon uses an example of the `Metro UI Colors Color Palette` from [color-hex.com](http://www.color-hex.com/color-palette/700) and
shows how to use named colors and how to write helper-functions to make
it easy to work with the `tidyverse`.

- **Slide design** by [Zach Holman](http://zachholman.com/posts/slide-design-for-developers/)

- Color palettes from [ColorBrewer](http://colorbrewer2.org/)    
**Note:** To find those suitable for those who are colorblind use

```{r findPalletesForColorblind}
subset(RColorBrewer::brewer.pal.info , colorblind)
```

Remember it can be used with `gglot2` via `scale_color_brewer()` and
`scale_fill_brewer()`.    
**Credit:**
[K. Vitting-Seerup](https://twitter.com/KVittingSeerup/status/1011898955477774336)
(2018-06-27).

- Color palettes from [colourlovers](http://www.colourlovers.com)

- Color palettes from [color-hex.com](http://www.color-hex.com/color-palettes/)

- Color palettes from [Kuler](https://kuler.adobe.com/create/color-wheel/)

- [efg's handy R-Color PDF](pdf/RColorChart.pdf)


# My favorite desktop background

I like ``RGB(0,34,64)`` which is ``#002240``. Note that my favorite
light yellow background for analySIS scale bars is ``RGB(255,255,220)``
which is ``#FFFFDC`` 

```{r,fig.height=1, fig.width=1}
pal <- c("#002240", "#FFFFDC")
std.mar <- par()$mar
par(mar = rep(0, 4)) 
image(matrix(1:2), col = pal, axes = FALSE)
par(mar=std.mar)
```


# Some Palettes I like

## Colors that look OK with Kodak colors...
```
Kodak trade dress yellow: #FFCC00
   Kodak trade dress red: #CC0000
        Third-party blue: #3300FF
                80% Gray: #2F2F2F
                60% Gray: #5F5F5F
                30% Gray: #B2B2B2
                20% Gray: #D2D2D2
                10% Gray: #EFEFEF
```

```{r,fig.height=1}
pal <- c("#FFCC00","#CC0000","#3300FF","#2F2F2F","#5F5F5F","#B2B2B2","#D2D2D2","#EFEFEF")
names(pal) <- c("","","","","","","","")
std.mar <- par()$mar
par(mar = rep(0, 4)) 
image(matrix(1:8), col = pal, axes = FALSE)
par(mar=std.mar)
```

or my own 7 step mix -- too garish...

```{r,fig.height=1}
# pal <- c("#CC0000","#D62900","#E05200","#EB7A00","#F5A300","#FFCC00","#D6A333","#AD7A66","#6633BF","#5C29CC","#3300FF")
pal <- c("#CC0000","#E05200","#F5A300","#FFCC00","#D6A333","#6633BF","#3300FF")
names(pal) <- c("","","","","","","")
std.mar <- par()$mar
par(mar = rep(0, 4)) 
image(matrix(1:7), col = pal, axes = FALSE)
par(mar=std.mar)
```

## 8-class BrBG (Diverging) from Color Brewer

```
Brown
#8c510a
#bf812d
#dfc27d
#f6e8c3 x
#c7eae5 x
#80cdc1
#35978f
#01665e
Blue-Green
```
```{r,fig.height=1}
pal <- c("#8c510a","#bf812d", "#dfc27d", "#f6e8c3", "#c7eae5","#80cdc1","#35978f","#01665e")
names(pal) <- c("","","","","","","","")
std.mar <- par()$mar
par(mar = rep(0, 4)) 
image(matrix(1:8), col = pal, axes = FALSE)
par(mar=std.mar)
```



## 8-class  RdYlBu

```
Red
#d73027
#f46d43
#fdae61
#fee090
#e0f3f8
#abd9e9
#74add1
#4575b4
Blue
```
```{r,fig.height=1}
pal <- c("#d73027","#f46d43","#fdae61","#fee090","#e0f3f8","#abd9e9","#74add1","#4575b4")
names(pal) <- c("","","","","","","","")
std.mar <- par()$mar
par(mar = rep(0, 4)) 
image(matrix(1:8), col = pal, axes = FALSE)
par(mar=std.mar)
```



[Business Casual](https://kuler.adobe.com/Business-Casual-color-theme-5595/)

```{r,fig.height=1}
pal <- c("#133463","#365FB7", "#5A72A6", "#F4EFDC", "#BA9B65")
std.mar <- par()$mar
par(mar = rep(0, 4)) 
image(matrix(1:5), col = pal, axes = FALSE)
par(mar=std.mar)
```


[Threadless](http://www.colourlovers.com/palette/27905/threadless)

```{r,fig.height=1}
pal <- c("#1B325F","#9CC4E4", "#E9F2F9", "#3A89C9", "#F26C4F")
std.mar <- par()$mar
par(mar = rep(0, 4)) 
image(matrix(1:5), col = pal, axes = FALSE)
par(mar=std.mar)
```

[Blown Entrepreneur](http://www.colourlovers.com/palette/59867/Blown_Entrepreneur)

```{r,fig.height=1}
pal <- c("#44749D","#C6D4E1", "#FFFFFF", "#EBE7E0", "#BDB8AD")
names(pal) <- c("","", "", "", "")
std.mar <- par()$mar
par(mar = rep(0, 4)) 
image(matrix(1:5), col = pal, axes = FALSE)
par(mar=std.mar)
```

[Back to Index](../README.html)
