---
title: "R Scrape PDF"
author: "J. R. Minter"
date: "Started: 2019-05-10, Last modified: 2019-05-10"
output:
  html_document:
    css: ../theme/jm-gray-vignette.css
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
library(knitr)
options(width = 72)

```


[Back to Index](../README.html)

# `tabulizer:` Scraping tables from PDF files

## An example from the vignette

```{r scrape_example, warning=FALSE}
library("tabulizer")
f <- system.file("examples", "data.pdf", package = "tabulizer")
out1 <- extract_tables(f)
str(out1)
```

Note that the output is a list of tables.

List the first table:

```{r list_table_1, comment=NA}
out1[[1]]
```

List the second table:

```{r list_table_2, comment=NA}
out1[[2]]

```

List the third table:


```{r list_table_3, comment=NA}
out1[[3]]
```

## The extract_areas() function

Tabulizer has a function, `extract_areas()` :

This could be **_really useful_**!

```
extract_areas(file, pages = NULL, guess = FALSE, copy = FALSE, ...)
```

| Argument | Description |
|----------|---------------|
| file | A character string specifying the path to a PDF file. This can also be a URL, in which case the file will be downloaded to the R temporary directory using download.file.|
| pages | An optional integer vector specifying pages to extract from. To extract multiple tables from a given page, repeat the page number (e.g., c(1,2,2,3)).|
| guess | See extract_tables (note the different default value).|
| copy | Specifies whether the original local file(s) should be copied to tempdir() before processing. FALSE by default. The argument is ignored if file is URL.|
| ... |Other arguments passed to extract_tables.|




[Back to Index](../README.html)
